\documentclass[12pt,a4paper]{article}

% ---- Packages ----
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,fit,backgrounds,calc}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage[round]{natbib}
\usepackage{fancyhdr}
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{float}
\usepackage{everypage}
\usepackage{eso-pic}
\usepackage{longtable}

% ---- GrokRxiv Sidebar ----
\definecolor{grokgray}{RGB}{110,110,110}
\AddEverypageHook{%
  \ifnum\value{page}=1
    \begin{tikzpicture}[remember picture, overlay]
      \node[
        rotate=90,
        anchor=south,
        font=\Large\sffamily\bfseries\color{grokgray},
        inner sep=0pt
      ] at ([xshift=38pt, yshift=0.52\paperheight]current page.south west)
      {GrokRxiv:2026.03.ai-os-synthesis\quad
       [\,cs.AI\,]\quad
       10 Mar 2026};
    \end{tikzpicture}
  \fi
}

% ---- Theorem Environments ----
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}

% ---- Colors ----
\definecolor{elixirpurple}{RGB}{75,0,130}
\definecolor{codegreen}{RGB}{0,128,0}
\definecolor{codegray}{RGB}{128,128,128}
\definecolor{codebg}{RGB}{248,248,248}

% ---- Listings ----
\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{elixirpurple}\bfseries,
  commentstyle=\color{codegreen},
  stringstyle=\color{red!60!black},
  numbers=left,
  numberstyle=\tiny\color{codegray},
  numbersep=8pt,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{codegray!40},
  breaklines=true,
  captionpos=b,
  tabsize=2,
  showstringspaces=false,
  xleftmargin=2em,
  framexleftmargin=1.5em
}

\lstdefinelanguage{elixir}{
  morekeywords={defmodule,def,defp,defstruct,do,end,use,require,alias,
    case,cond,fn,if,else,when,with,for,true,false,nil,
    @moduledoc,@doc,@spec,@type,@typedoc,@impl,@callback,
    @enforce_keys,@behaviour,@default_timeout},
  sensitive=true,
  morecomment=[l]{\#},
  morestring=[b]",
  morestring=[b]'
}

% ---- Header/Footer ----
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textit{The AI Operating System: Part V --- Modular Synthesis}}
\fancyhead[R]{\small\thepage}
\fancyfoot[C]{\small The YonedaAI Collaboration}

% ---- Title ----
\title{%
  \textbf{The AI Operating System}\\[0.5em]
  \Large Part V: Modular Synthesis ---\\
  Composing Four Subsystems into a Unified Runtime\\[1em]
  \large An Elixir/OTP Umbrella Application
}

\author{%
  \textbf{Matthew Long}\\
  The YonedaAI Collaboration\\
  YonedaAI Research Collective\\
  Chicago, IL\\
  \texttt{matthew@yoneda.ai}
}

\date{March 10, 2026}

\begin{document}

\maketitle

% ========================================================================
\begin{abstract}
Parts I--IV of \textit{The AI Operating System} developed four independent subsystems: an
\textsc{Agent Scheduler} for process management, a \textsc{Tool Interface}
for capability-based access control, a \textsc{Memory Layer} for typed persistent
cognition, and a \textsc{Planner Engine} for market-based orchestration.
Each subsystem was designed as a self-contained Elixir/OTP application with its own
supervision tree, public API, and type specifications.

This paper addresses the synthesis question: \emph{how do these four modules compose
into a single, coherent operating system for AI agents?}

We present \texttt{AgentOS}, an Elixir umbrella application that composes the four
subsystems through (1)~a dependency-ordered supervision tree,
(2)~message-passing interfaces between subsystem boundaries,
(3)~pipeline composition across layers, and (4)~a unified facade module that
exposes the complete job lifecycle.
We show that every pairwise combination of subsystems yields emergent capabilities:
agents invoke tools via capabilities, agents persist state through memory,
tool results are cached in typed memory, and the planner orchestrates all three
through market mechanisms.
The result is a production-grade system demonstrated through the Agent-Hero
marketplace---a Next.js + Supabase + Stripe platform where clients post jobs,
the planner decomposes and matches them to agents via an order book, agents
execute with tool access and memory persistence, and evaluators score outcomes
to update reputation.

The implementation comprises 24 Elixir source files totaling approximately 3{,}500
lines of code, running on the BEAM virtual machine.
All composition is achieved through standard functional programming
patterns: module boundaries, supervision hierarchies, typed message passing,
behaviour callbacks, and pipeline composition.
\end{abstract}

\vspace{1em}
\noindent\textbf{Keywords:} AI operating systems, modular composition, Elixir/OTP, agent orchestration,
supervision trees, capability-based security, typed memory, market mechanisms

\clearpage
\tableofcontents
\clearpage

% ========================================================================
\section{Introduction}\label{sec:intro}

The preceding four papers in this series each addressed a single concern
of an AI operating system:

\begin{enumerate}[label=\textbf{Part \Roman*:}, leftmargin=3.5em]
  \item \textbf{Agent Scheduler} --- Process lifecycle management for AI agents,
        including credit-weighted scheduling, durable execution with memoization,
        streaming pipelines, and 6-dimensional evaluation.
  \item \textbf{Tool Interface} --- Capability-based access control with a 3-tier
        registry (builtin, sandbox, MCP), sandboxed execution via isolated BEAM
        processes, and per-invocation audit logging.
  \item \textbf{Memory Layer} --- A typed filesystem for persistent agent cognition,
        with 24 schema types across 6 subcategories, versioned evolution with vector
        clocks, graph relationships, and multi-backend storage.
  \item \textbf{Planner Engine} --- Market-based orchestration with an order book
        matching engine, escrow-backed financial transactions, DAG-based job
        decomposition, and reputation scoring with anti-gaming detection.
\end{enumerate}

Each part was designed as an independent Elixir/OTP application---a self-contained
module with its own supervision tree, public API, type specifications, and test suite.
This modularity is not accidental; it is the central architectural principle.

\subsection{The Synthesis Problem}

The challenge addressed by this paper is \emph{composition}: given four well-defined
modules, how do we assemble them into a system that is greater than the sum of its parts?
Specifically:

\begin{itemize}
  \item What is the correct startup ordering, and why does it matter for fault tolerance?
  \item How do subsystems communicate across module boundaries without violating encapsulation?
  \item What emergent capabilities arise from pairwise and higher-order compositions?
  \item How does the composed system support a complete production job lifecycle?
\end{itemize}

\subsection{Approach: Functional Composition on the BEAM}

Our approach is grounded in functional programming on the BEAM virtual machine.
The composition mechanisms are:

\begin{enumerate}
  \item \textbf{Supervision trees} --- OTP supervisors provide fault-isolated startup
        ordering with automatic restart on failure.
  \item \textbf{Message passing} --- Subsystems communicate via \texttt{GenServer.call/cast}
        across process boundaries, with no shared mutable state.
  \item \textbf{Behaviour callbacks} --- Elixir behaviours define type-safe interfaces
        that subsystems implement, enabling polymorphic dispatch.
  \item \textbf{Pipeline composition} --- Multi-stage streaming pipelines compose
        agents, tools, and memory into end-to-end execution flows.
  \item \textbf{Module facades} --- A top-level \texttt{AgentOS} module provides a
        unified API that delegates to the appropriate subsystem.
\end{enumerate}

None of these mechanisms require exotic abstractions.
They are the standard tools of Erlang/OTP, refined over 40 years of production
use in telecommunications, banking, and messaging infrastructure.

\subsection{Mathematical Foundations}\label{subsec:math-foundations}

The design of AgentOS is informed by ideas from category theory, which provides
a language for describing composition, interfaces, and structure preservation.
In particular:

\begin{itemize}
  \item Each subsystem can be understood as an object in a module category, with
        its public API defining the morphisms available to other subsystems.
  \item Composition of subsystems corresponds to constructing products and coproducts
        in this category: the umbrella application is a product of its component
        applications, while the storage router is a coproduct of backends.
  \item The type-safe interfaces between subsystems (behaviours, typespecs, message
        protocols) act as natural transformations that preserve structure across
        module boundaries.
\end{itemize}

However, the implementation does not require the reader to be fluent in categorical
language. Throughout this paper, we present all constructions concretely in terms of
Elixir modules, supervision trees, and message-passing protocols. The categorical
perspective serves as background context for readers who find it illuminating, not
as a prerequisite for understanding the system.

\subsection{Paper Outline}

Section~\ref{sec:subsystems} reviews the four subsystems and their public interfaces.
Section~\ref{sec:umbrella} presents the umbrella application and its supervision tree.
Section~\ref{sec:pairwise} analyzes the six pairwise compositions and their emergent
capabilities.  Section~\ref{sec:full-composition} shows the full four-way composition
through a complete job lifecycle.  Section~\ref{sec:production} describes the
production deployment in the Agent-Hero marketplace.  Section~\ref{sec:properties}
establishes key system properties: fault isolation, type safety, and resource fairness.
Section~\ref{sec:related} surveys related work, and Section~\ref{sec:conclusion}
concludes.


% ========================================================================
\section{The Four Subsystems}\label{sec:subsystems}

Before examining their composition, we summarize each subsystem's architecture,
supervision tree, and public API. The goal is to establish the \emph{interfaces}
through which composition occurs.

% ------------------------------------------------------------------------
\subsection{Agent Scheduler}\label{subsec:scheduler}

The Agent Scheduler manages the lifecycle of AI agents as supervised BEAM processes.
Its architecture mirrors classical OS process management, adapted for AI workloads.

\subsubsection{Supervision Tree}

\begin{lstlisting}[language=elixir, caption={Agent Scheduler supervision tree}]
AgentScheduler.AppSupervisor (rest_for_one)
  +-- Registry (unique, :agent_registry)
  +-- AgentScheduler.Evaluator (GenServer)
  +-- AgentScheduler.Scheduler (GenServer)
  +-- AgentScheduler.Pipeline (GenServer)
  +-- AgentScheduler.Supervisor (DynamicSupervisor)
      +-- Agent "agent_001"
      +-- Agent "agent_002"
      +-- Agent "agent_n"
\end{lstlisting}

The \texttt{:rest\_for\_one} strategy ensures that if the Registry crashes,
all dependent components restart in order.
The DynamicSupervisor uses \texttt{:one\_for\_one} so individual agent failures
are isolated.

\subsubsection{Key Modules}

\begin{description}[style=nextline]
  \item[\texttt{AgentScheduler.Agent}]
    A GenServer managing an individual agent's state machine:
    \texttt{:pending} $\to$ \texttt{:running} $\to$ \texttt{:waiting\_approval}
    $\to$ \texttt{:completed}.
    Implements durable execution via a memoization store: each step function is
    keyed by a unique identifier, and completed steps are replayed from the memo
    store on crash recovery without re-execution.

  \item[\texttt{AgentScheduler.Scheduler}]
    A credit-weighted priority scheduler using Erlang's \texttt{:gb\_trees}
    (balanced binary trees) for $O(\log n)$ enqueue/dequeue.
    Each client accumulates \emph{agent virtual runtime} (avruntime)
    proportional to credits consumed:
    \begin{equation}\label{eq:avruntime}
      \text{avruntime}(c) \mathrel{+}= \frac{\kappa_A \cdot \text{cr}_0}
                                              {\text{cr}_c \cdot w(p_c)}
    \end{equation}
    where $\kappa_A$ is agent cost, $\text{cr}_0 = 1000$ is the reference credit
    amount, $\text{cr}_c$ is the client's remaining credits, and $w(p_c)$ is the
    priority weight (4.0 for contracted, 1.0 for marketplace).
    The scheduler always dispatches the client with the smallest avruntime.

  \item[\texttt{AgentScheduler.Pipeline}]
    Streaming pipelines with EventEmitter-style pub/sub. Each pipeline consists of
    ordered stages declaring what event types they publish and subscribe to. The
    pipeline constraint ensures stages only subscribe to events from earlier stages:
    $\text{sub}(s_i) \subseteq \bigcup_{j < i} \text{pub}(s_j)$.

  \item[\texttt{AgentScheduler.Evaluator}]
    Six-dimensional quality evaluation across: quality, adherence, speed, cost
    efficiency, error rate, and revision count. The composite score is a weighted
    inner product, and reputation is computed as an EWMA:
    $R_t = \alpha \cdot s_t + (1 - \alpha) \cdot R_{t-1}$ with $\alpha = 0.3$.
\end{description}

\subsubsection{Public Interface}

\begin{lstlisting}[language=elixir, caption={Agent Scheduler public API}]
AgentScheduler.submit_job(client_id, job, opts)
  :: {:ok, job_id} | {:error, term()}

AgentScheduler.start_agent(agent_id, profile, opts)
  :: {:ok, pid()} | {:error, term()}

AgentScheduler.create_pipeline(name, stages)
  :: {:ok, pipeline_id} | {:error, term()}

AgentScheduler.get_evaluation(agent_id)
  :: {:ok, map()} | {:error, :not_found}
\end{lstlisting}

% ------------------------------------------------------------------------
\subsection{Tool Interface}\label{subsec:tools}

The Tool Interface mediates between agents and external systems, analogous to
device drivers in a classical operating system.

\subsubsection{Three-Tier Registry}

Tools are organized into three trust tiers:

\begin{table}[H]
\centering
\caption{Tool tiers in the 3-tier registry}
\begin{tabular}{lll}
\toprule
\textbf{Tier} & \textbf{Trust Level} & \textbf{Examples} \\
\midrule
Builtin  & Static, trusted   & \texttt{web-search}, \texttt{text-transform}, \texttt{pdf-parse} \\
Sandbox  & Isolated execution & \texttt{code-exec}, \texttt{shell-exec}, \texttt{git-clone} \\
MCP      & Runtime-discovered & \texttt{server\_\_tool} (namespaced) \\
\bottomrule
\end{tabular}
\end{table}

The registry supports \emph{freezing}: once frozen, no new tools can be registered.
This implements contract-locked configuration where the tool set is immutable during
agent execution.

\subsubsection{Capability Tokens}

Access control is capability-based. Each agent holds unforgeable tokens that
grant specific permissions on specific tools:

\begin{lstlisting}[language=elixir, caption={Capability token structure}]
%ToolInterface.Capability{
  agent_id: "agent-1",
  tool_id: "web-search",
  permissions: [:invoke, :inspect],
  rate_limit: 60,           # invocations per minute
  expires_at: ~U[...],
  signature: <<hmac_sha256>>  # tamper detection
}
\end{lstlisting}

Authorization checks: (1)~tool ID matches, (2)~token not expired,
(3)~HMAC signature valid, (4)~\texttt{:invoke} permission present,
(5)~rate limit not exceeded. Constant-time comparison prevents timing attacks.

\subsubsection{Sandboxed Execution}

Sandbox-tier tools execute in isolated BEAM processes with their own heaps:

\begin{lstlisting}[language=elixir, caption={Sandbox execution via spawn\_monitor}]
def execute(tool_spec, input, opts) do
  timeout = Keyword.get(opts, :timeout, 600_000)
  parent = self()
  ref = make_ref()

  {pid, monitor_ref} = spawn_monitor(fn ->
    result = tool_spec.execute.(input)
    send(parent, {:sandbox_result, ref, result})
  end)

  receive do
    {:sandbox_result, ^ref, result} ->
      Process.demonitor(monitor_ref, [:flush])
      result
    {:DOWN, ^monitor_ref, :process, ^pid, reason} ->
      {:error, {:sandbox_crashed, reason}}
  after
    timeout ->
      Process.exit(pid, :kill)
      {:error, :timeout}
  end
end
\end{lstlisting}

Each execution is time-bounded (default 10 minutes, matching E2B sandbox behavior),
error-contained, and cleaned up via monitor-based process reaping.

\subsubsection{Audit Logging}

Every tool invocation is logged asynchronously via \texttt{GenServer.cast/2},
recording: agent ID, tool ID, input (with sensitive fields redacted), output,
duration, and status. The asynchronous cast ensures audit overhead does not
impact tool invocation latency.

\subsubsection{Public Interface}

\begin{lstlisting}[language=elixir, caption={Tool Interface public API}]
ToolInterface.invoke(agent_id, tool_id, input, capability_token)
  :: {:ok, result} | {:error, term()}

ToolInterface.grant_capability(agent_id, tool_id, opts)
  :: {:ok, Capability.t()} | {:error, term()}

ToolInterface.discover_mcp(server_name, url, token)
  :: :ok | {:error, term()}

ToolInterface.freeze()
  :: :ok

ToolInterface.list_tools(tier)
  :: [tool_spec()]
\end{lstlisting}

% ------------------------------------------------------------------------
\subsection{Memory Layer}\label{subsec:memory}

The Memory Layer provides a typed filesystem for persistent agent cognition,
with 24 schema types organized into 6 subcategories.

\subsubsection{Schema Type System}

\begin{table}[H]
\centering
\caption{The 24 memory types across 6 subcategories}
\begin{tabular}{ll}
\toprule
\textbf{Subcategory} & \textbf{Types} \\
\midrule
Core       & \texttt{fact}, \texttt{decision}, \texttt{procedural}, \texttt{episodic} \\
Extended   & \texttt{todo}, \texttt{issue}, \texttt{api}, \texttt{schema\_def} \\
Workflow   & \texttt{workflow}, \texttt{task}, \texttt{step}, \texttt{agent\_run} \\
Metadata   & \texttt{tag}, \texttt{annotation}, \texttt{embedding}, \texttt{index} \\
System     & \texttt{config}, \texttt{session}, \texttt{context}, \texttt{log} \\
Relational & \texttt{person}, \texttt{project}, \texttt{artifact}, \texttt{event} \\
\bottomrule
\end{tabular}
\end{table}

Each schema is an Elixir struct with \texttt{@enforce\_keys} for compile-time
safety and \texttt{@type} annotations for Dialyzer checking. The Schema Registry
maps name strings to modules for runtime type resolution:

\begin{lstlisting}[language=elixir, caption={Schema Registry: name to module resolution}]
# Initialization
%{
  "fact"       => MemoryLayer.Schema.FactData,
  "decision"   => MemoryLayer.Schema.DecisionData,
  "procedural" => MemoryLayer.Schema.ProceduralData,
  "episodic"   => MemoryLayer.Schema.EpisodicData,
  "todo"       => MemoryLayer.Schema.TodoData,
  "issue"      => MemoryLayer.Schema.IssueData,
  "workflow"   => MemoryLayer.Schema.WorkflowData,
  "agent_run"  => MemoryLayer.Schema.AgentRunData
}
\end{lstlisting}

\subsubsection{Dual-Layer Storage}

Memories are persisted to two backends simultaneously:

\begin{description}
  \item[ETS (Working Memory)] --- In-process, microsecond access, volatile.
    Analogous to human working memory. Used for active memories during
    agent execution.
  \item[Mnesia (Persistent Knowledge)] --- Distributed, ACID-transactional,
    disk-backed. Analogous to long-term memory. Survives process crashes
    and node restarts.
\end{description}

The storage router implements a read-through cache: recall checks ETS first
(fast path), falls back to Mnesia (durable path), and promotes retrieved
memories to ETS for future access. LRU eviction keeps working memory bounded.

\subsubsection{Versioning and Graph}

Memories evolve over time. The \texttt{evolve/3} operation creates a new
memory linked to its parent via:

\begin{enumerate}
  \item A \textbf{version record} in Mnesia with causal annotation
        (observation, inference, correction, decay) and a vector clock
        for multi-agent conflict detection.
  \item A \textbf{graph edge} connecting parent to child via one of nine
        typed relations: \texttt{evolved\_into}, \texttt{references},
        \texttt{supersedes}, \texttt{contradicts}, \texttt{supports},
        \texttt{derived\_from}, \texttt{part\_of}, \texttt{triggers},
        \texttt{blocked\_by}.
\end{enumerate}

Content hashing (SHA-256) enables deduplication: two memories with
identical content share the same hash.

\subsubsection{Public Interface}

\begin{lstlisting}[language=elixir, caption={Memory Layer public API}]
MemoryLayer.Memory.create(schema_struct)
  :: {:ok, pid()} | {:error, term()}

MemoryLayer.Memory.data(pid)
  :: {:ok, struct()} | {:error, :deleted}

MemoryLayer.Memory.evolve(pid, changes, reason)
  :: {:ok, child_pid} | {:error, term()}

MemoryLayer.Storage.search(query, opts)
  :: {:ok, [map()]} | {:error, term()}

MemoryLayer.Graph.link(from_id, to_id, relation, metadata)
  :: :ok | {:error, term()}

MemoryLayer.Graph.traverse(start_id, pattern, opts)
  :: [String.t()]
\end{lstlisting}

% ------------------------------------------------------------------------
\subsection{Planner Engine}\label{subsec:planner}

The Planner Engine is the highest-level orchestration layer, implementing
market-based job matching and financial settlement.

\subsubsection{Order Book}

The order book maintains two sides:

\begin{description}
  \item[Demands (Buy Orders)] --- Posted by clients, specifying required
    capabilities, budget ceiling, and deadline.
  \item[Proposals (Sell Orders)] --- Submitted by agents, specifying
    execution plan, estimated credits, duration, and confidence score.
\end{description}

Matching uses price-time priority: proposals are sorted by
\texttt{(estimated\_credits ASC, timestamp ASC)}. A match occurs when the
best proposal's credits $\leq$ demand's budget ceiling. Upon match, all
other proposals are auto-rejected.

The cost functional for ranking proposals is:
\begin{equation}\label{eq:cost-functional}
  \text{cost}(\alpha, \tau) = \frac{\text{estimated\_credits}}
                                    {1 + \text{confidence} \times \text{reputation}}
\end{equation}

\subsubsection{Escrow}

Financial transactions use Mnesia-backed escrow:

\begin{enumerate}
  \item \texttt{hold(client\_id, amount, contract\_id)} --- Atomically moves
        credits from available to held. Fails on insufficient funds.
  \item \texttt{settle(escrow\_id, :release)} --- Transfers held credits to
        the operator (job completed successfully).
  \item \texttt{settle(escrow\_id, :refund)} --- Returns held credits to the
        client (job cancelled or disputed).
\end{enumerate}

All balance mutations occur within Mnesia transactions, providing atomicity
(no TOCTOU races), isolation (serialized concurrent operations), and durability
(committed transactions survive crashes).

The \textbf{conservation invariant} is maintained: total credits in the system
(available + held + distributed) is constant. No operation creates or destroys credits.

\subsubsection{Job Decomposition}

The Decomposer transforms a task into a DAG of subtasks with dependency edges.
Topological sort produces execution levels where each level contains subtasks
that can execute in parallel:

\begin{lstlisting}[language=elixir, caption={Parallel execution schedule from DAG decomposition}]
task = %{
  id: "task_42",
  subtask_specs: [
    %{description: "E2E tests",
      required_capabilities: [:playwright],
      estimated_credits: 1000, estimated_duration: 60},
    %{description: "Load tests",
      required_capabilities: [:k6],
      estimated_credits: 800, estimated_duration: 45},
    %{description: "Generate report",
      required_capabilities: [:reporting],
      estimated_credits: 200, estimated_duration: 15,
      depends_on: [0, 1]}
  ]
}

{:ok, dag} = PlannerEngine.Decomposer.decompose(task)
schedule = PlannerEngine.Decomposer.topological_sort(dag)
# => [["task_42_sub_0", "task_42_sub_1"], ["task_42_sub_2"]]
#     Level 0: E2E + Load in parallel
#     Level 1: Report after both complete
\end{lstlisting}

\subsubsection{Revenue Distribution}

Upon contract completion, escrowed credits are distributed:

\begin{table}[H]
\centering
\caption{Revenue split on contract completion}
\begin{tabular}{lr}
\toprule
\textbf{Recipient} & \textbf{Share} \\
\midrule
Operator (agent who performed work) & 70\% \\
Platform (marketplace fee)          & 15\% \\
LLM Reserve (inference costs)       & 15\% \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Reputation and Anti-Gaming}

The reputation system scores agents across 6 dimensions (accuracy, completeness,
timeliness, communication, efficiency, innovation) using exponentially-weighted
moving averages with decay factor $\lambda = 0.95$:

\begin{equation}\label{eq:reputation}
  R(\alpha) = \frac{\sum_{i=1}^{N} \lambda^{N-i} \langle w, q_i \rangle}
                    {\sum_{i=1}^{N} \lambda^{N-i}}
\end{equation}

The trust score incorporates anti-gaming detection:
\begin{equation}\label{eq:trust}
  T(\alpha) = R(\alpha) \times (1 - \gamma(\alpha)) \times \min\!\Big(1, \frac{N(\alpha)}{N_{\min}}\Big)
\end{equation}
where $\gamma(\alpha)$ is the gaming suspicion score (0 to 1) and $N_{\min} = 5$
is the seasoning threshold. Five anomaly detectors run on each evaluation:
score inflation, rapid cycling, perfect streaks, self-dealing, and collusion rings.


% ========================================================================
\section{The Umbrella Application}\label{sec:umbrella}

With the four subsystems defined, we now construct the \texttt{AgentOS}
umbrella application that composes them into a unified runtime.

\subsection{Mix Project Configuration}

In Elixir's build system, an umbrella application declares its child
applications as path dependencies:

\begin{lstlisting}[language=elixir, caption={mix.exs for the AgentOS umbrella}]
defmodule AgentOS.MixProject do
  use Mix.Project

  def project do
    [
      app: :agent_os,
      version: "0.1.0",
      elixir: "~> 1.17",
      deps: deps()
    ]
  end

  def application do
    [
      extra_applications: [:logger, :mnesia],
      mod: {AgentOS.Application, []}
    ]
  end

  defp deps do
    [
      {:agent_scheduler, path: "../agent_scheduler"},
      {:tool_interface, path: "../tool_interface"},
      {:memory_layer, path: "../memory_layer"},
      {:planner_engine, path: "../planner_engine"},
      {:jason, "~> 1.4"},
      {:telemetry, "~> 1.3"}
    ]
  end
end
\end{lstlisting}

The \texttt{path:} dependencies tell Mix that these are local applications
within the same repository. Each compiles independently and exposes its
public modules to the umbrella.

\subsection{Startup Ordering}\label{subsec:startup}

The most critical design decision in the umbrella is the \emph{startup order}
of child applications. The order is not arbitrary---it reflects the dependency
graph between subsystems.

\begin{lstlisting}[language=elixir, caption={AgentOS.Application: ordered supervision tree}]
defmodule AgentOS.Application do
  use Application

  @impl true
  def start(_type, _args) do
    children = [
      # Layer 1: Memory -- foundation for all state
      {MemoryLayer, []},

      # Layer 2: Tools -- depends on memory for
      #          capability storage
      {ToolInterface, []},

      # Layer 3: Scheduler -- depends on tools and memory
      {AgentScheduler, []},

      # Layer 4: Planner -- depends on all three
      {PlannerEngine, []}
    ]

    opts = [strategy: :one_for_one,
            name: AgentOS.Supervisor]
    Supervisor.start_link(children, opts)
  end
end
\end{lstlisting}

\subsubsection{Why This Order?}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=1.8cm,
  layer/.style={draw, rounded corners, minimum width=10cm,
                minimum height=1.2cm, font=\ttfamily\small,
                fill=blue!5},
  arrow/.style={-{Stealth[length=8pt]}, thick, gray}
]
  \node[layer, fill=purple!8] (planner) {Layer 4: PlannerEngine};
  \node[layer, fill=green!8, below=0.5cm of planner] (scheduler) {Layer 3: AgentScheduler};
  \node[layer, fill=orange!8, below=0.5cm of scheduler] (tools) {Layer 2: ToolInterface};
  \node[layer, fill=blue!8, below=0.5cm of tools] (memory) {Layer 1: MemoryLayer};

  \draw[arrow] (planner.south) -- (scheduler.north) node[midway, right, font=\scriptsize] {orchestrates agents};
  \draw[arrow] (scheduler.south) -- (tools.north) node[midway, right, font=\scriptsize] {invokes tools};
  \draw[arrow] (tools.south) -- (memory.north) node[midway, right, font=\scriptsize] {stores capabilities};
  \draw[arrow] (planner.west) to[out=180,in=180] node[midway, left, font=\scriptsize] {reads reputation} (memory.west);
  \draw[arrow] (scheduler.west) to[out=190,in=170] node[midway, left, font=\scriptsize\hspace{-1em}] {persists state} (memory.west);
\end{tikzpicture}
\caption{AgentOS supervision tree with dependency arrows. Each layer depends on
all layers below it. Startup proceeds bottom-up.}
\label{fig:supervision}
\end{figure}

The ordering reflects a strict dependency hierarchy:

\begin{enumerate}
  \item \textbf{MemoryLayer starts first} because it initializes ETS tables
        (\texttt{:memory\_working}, \texttt{:memory\_lru}) and Mnesia tables
        (\texttt{:memories}, \texttt{:versions}, \texttt{:edges}) that all other
        subsystems depend on. If memory is unavailable, nothing can persist state.

  \item \textbf{ToolInterface starts second} because it needs memory available for
        storing capability tokens and audit logs. The tool registry itself
        (builtin + sandbox tools) is initialized at startup.

  \item \textbf{AgentScheduler starts third} because agents need both tools
        (to execute steps) and memory (to persist memoization stores and checkpoints).
        The Evaluator, Scheduler, Pipeline, and DynamicSupervisor all start in this phase.

  \item \textbf{PlannerEngine starts last} because it orchestrates all three lower
        layers. The order book needs the scheduler to dispatch matched agents.
        The escrow system needs Mnesia (initialized by MemoryLayer). The reputation
        system needs the evaluator to provide quality scores.
\end{enumerate}

\subsubsection{The one\_for\_one Strategy}

The top-level supervisor uses \texttt{:one\_for\_one}: if any single subsystem
crashes, only that subsystem is restarted. The other three continue operating.
This is appropriate because each subsystem manages its own state independently
via GenServers. A crash in the Planner Engine does not corrupt the Memory Layer's
ETS tables, and vice versa.

If the system required that a Planner crash also restart the Scheduler (because
the Planner holds references to scheduler state), we would use
\texttt{:rest\_for\_one}. The current design avoids this coupling.

\subsection{The Unified Facade}\label{subsec:facade}

The \texttt{AgentOS} module provides a simplified entry point that delegates
to the appropriate subsystem:

\begin{lstlisting}[language=elixir, caption={AgentOS facade module}]
defmodule AgentOS do
  @spec start(keyword()) :: {:ok, pid()} | {:error, term()}
  def start(opts \\ []) do
    AgentOS.Application.start(:normal, opts)
  end

  @spec submit_job(map()) :: {:ok, String.t()} | {:error, term()}
  def submit_job(job_spec) do
    PlannerEngine.OrderBook.post_demand(job_spec)
  end

  @spec status() :: map()
  def status do
    %{
      scheduler: %{running: true},
      tools: ToolInterface.list_tools(),
      memory: %{running: true},
      planner: %{running: true}
    }
  end
end
\end{lstlisting}

Clients interact with \texttt{AgentOS.submit\_job/1} and never need to know
which subsystem handles their request. The facade pattern keeps the public
API surface small while the internal implementation spans four applications.


% ========================================================================
\section{Pairwise Compositions}\label{sec:pairwise}

The four subsystems yield $\binom{4}{2} = 6$ pairwise compositions.
Each pair produces emergent capabilities that neither subsystem provides alone.

\subsection{Agent Scheduler + Tool Interface}\label{subsec:agent-tool}

\emph{Emergent capability: agents invoke tools via capability tokens.}

When an agent is assigned a job, it receives capability tokens for the tools
required by that job. The composition works as follows:

\begin{lstlisting}[language=elixir, caption={Agent invoking a tool via capability}]
# 1. Start an agent with tool capabilities
{:ok, _pid} = AgentScheduler.start_agent("agent_1", %{
  name: "WebTester",
  capabilities: [:playwright, :k6],
  task_domain: [:web_testing]
})

# 2. Grant capability tokens for required tools
{:ok, cap_playwright} = ToolInterface.grant_capability(
  "agent_1", "code-exec",
  permissions: [:invoke], rate_limit: 30
)

# 3. Agent executes a durable step that invokes a tool
AgentScheduler.Agent.execute_step("agent_1", "run_e2e", fn ->
  ToolInterface.invoke("agent_1", "code-exec",
    %{"code" => "playwright_script", "language" => "javascript"},
    cap_playwright)
end)
\end{lstlisting}

The key insight is that \texttt{execute\_step/3} wraps the tool invocation
in durable execution: if the agent crashes after the tool returns but before
the step is recorded, the memoization store replays the cached result on restart
instead of re-invoking the tool.

\textbf{Type safety at the boundary:}
The capability token's \texttt{tool\_id} must match the tool being invoked.
The tool's \texttt{input\_schema} validates the input map. The agent's
\texttt{execute\_step} wrapper ensures the result is memoized with a unique
step ID. All three checks happen at different layers, composing into a
defense-in-depth guarantee.

\subsection{Agent Scheduler + Memory Layer}\label{subsec:agent-memory}

\emph{Emergent capability: agents persist state across executions.}

Agent memoization stores are in-process (they live in the GenServer's state).
When an agent crashes and restarts, the memo store is empty. By composing with
the Memory Layer, agents gain persistent state:

\begin{lstlisting}[language=elixir, caption={Agent persisting state to typed memory}]
# Agent creates a typed memory for its execution trace
{:ok, trace_pid} = MemoryLayer.Memory.create(
  %MemoryLayer.Schema.AgentRunData{
    agent_id: "agent_1",
    task: "web_testing",
    status: :running,
    started_at: DateTime.utc_now()
  })

# Each step result is persisted as a fact
AgentScheduler.Agent.execute_step("agent_1", "discover_pages", fn ->
  pages = discover_pages("https://example.com")

  {:ok, _} = MemoryLayer.Memory.create(
    %MemoryLayer.Schema.FactData{
      assertion: "Discovered #{length(pages)} pages",
      confidence: 1.0,
      source: "agent_1/discover_pages"
    })

  pages
end)
\end{lstlisting}

The Memory Layer's version tracking provides a complete audit trail: each
step produces a new memory linked to its predecessors via \texttt{:derived\_from}
edges. If the agent crashes and restarts, it can query the Memory Layer
to reconstruct its progress:

\begin{lstlisting}[language=elixir, caption={Agent recovering state from memory after crash}]
# After restart, query memory for previous results
{:ok, previous_results} = MemoryLayer.Storage.search(
  "agent_1",
  filter: %{schema_name: "agent_run"},
  backend: :mnesia  # Durable storage survives crashes
)
\end{lstlisting}

\subsection{Tool Interface + Memory Layer}\label{subsec:tool-memory}

\emph{Emergent capability: tool results are cached in typed memory.}

Tool invocations can be expensive (API calls, code execution, web scraping).
By composing tools with memory, results are cached and deduplicated:

\begin{lstlisting}[language=elixir, caption={Tool result caching via memory}]
def invoke_with_cache(agent_id, tool_id, input, cap) do
  # Check if result is already cached
  cache_key = MemoryLayer.Memory.content_hash(
    %{tool_id: tool_id, input: input})

  case MemoryLayer.Storage.search(cache_key, backend: :ets) do
    {:ok, [cached | _]} ->
      {:ok, cached.data}

    {:ok, []} ->
      # Cache miss: invoke tool, store result
      case ToolInterface.invoke(agent_id, tool_id, input, cap) do
        {:ok, result} ->
          {:ok, _} = MemoryLayer.Memory.create(
            %MemoryLayer.Schema.FactData{
              assertion: "Tool result for #{tool_id}",
              confidence: 1.0,
              source: "tool_cache/#{cache_key}"
            })
          {:ok, result}

        error -> error
      end
  end
end
\end{lstlisting}

The SHA-256 content hashing from the Memory Layer enables exact deduplication:
identical inputs to the same tool always produce the same cache key.

Additionally, the audit log from \texttt{ToolInterface.Audit} can be
stored as \texttt{EpisodicData} memories, creating a queryable history of
all tool invocations with graph relationships to the agents that invoked them.

\subsection{Planner Engine + Agent Scheduler}\label{subsec:planner-scheduler}

\emph{Emergent capability: market-matched agents are automatically dispatched.}

When the Planner Engine clears the market for a task, it needs to assign the
matched agent to actually execute the work. This is the Planner-to-Scheduler
composition:

\begin{lstlisting}[language=elixir, caption={Market clearing dispatches to scheduler}]
# Market clears: best proposal accepted, escrow held
{:ok, contract} = PlannerEngine.Market.clear_market("task_42")

# Scheduler assigns the matched agent
:ok = AgentScheduler.Agent.assign_job(
  contract.operator_id,
  %{
    id: contract.task_id,
    task: :web_testing,
    input: %{url: "https://example.com"},
    contract_id: contract.id
  }
)

# Agent transitions: :pending -> :running
# Scheduler dispatches based on avruntime (Eq. 1)
\end{lstlisting}

The credit-weighted scheduling ensures fairness: clients who have consumed
fewer resources relative to their credit balance get priority.
Contracted clients preempt marketplace clients (two-tier scheduling via
the \texttt{tier} field in the priority key).

\subsection{Planner Engine + Memory Layer}\label{subsec:planner-memory}

\emph{Emergent capability: market state and reputation persist across restarts.}

The Planner Engine's Escrow system already uses Mnesia (initialized by the
Memory Layer) for atomic financial transactions. The reputation system
stores quality vectors that feed back into the order book's cost functional
(Equation~\ref{eq:cost-functional}).

When the Planner Engine restarts after a crash, Mnesia tables persist:
\begin{itemize}
  \item \texttt{:balances} --- Client credit balances (available + held)
  \item \texttt{:escrows} --- Active escrow records with status
\end{itemize}

Reputation histories can be stored as typed memories:

\begin{lstlisting}[language=elixir, caption={Reputation stored as typed memory}]
# After each evaluation, persist as memory
{:ok, _} = MemoryLayer.Memory.create(
  %MemoryLayer.Schema.EpisodicData{
    event: "quality_evaluation",
    outcome: :completed,
    context: %{
      agent_id: "agent_1",
      quality_vector: [0.87, 0.92, 0.78, 0.85, 0.95, 0.88],
      trust_score: 0.82
    },
    participants: ["agent_1", "client_1"]
  })
\end{lstlisting}

This creates a queryable history of all evaluations, with graph edges
connecting evaluations to the agents and contracts they reference.

\subsection{Planner Engine + Tool Interface}\label{subsec:planner-tool}

\emph{Emergent capability: the planner validates agent capabilities against tool requirements.}

When matching proposals to demands, the planner needs to verify that an
agent's claimed capabilities correspond to actual registered tools:

\begin{lstlisting}[language=elixir, caption={Capability verification during matching}]
def verify_capabilities(proposal, demand) do
  required = demand.required_capabilities
  registered_tools = ToolInterface.list_tools()
  tool_names = Enum.map(registered_tools, & &1.name)

  # Check that all required capabilities map to real tools
  Enum.all?(required, fn cap ->
    Atom.to_string(cap) in tool_names or
    Enum.any?(tool_names, &String.contains?(&1, Atom.to_string(cap)))
  end)
end
\end{lstlisting}

The tool registry freeze mechanism integrates with contract creation:
once a market clears and a contract is created, the tool registry is frozen
to ensure the tool set cannot change during execution.


% ========================================================================
\section{Full Four-Way Composition}\label{sec:full-composition}

We now trace a complete job through the fully composed system, showing how
all four subsystems interact.

\subsection{The Complete Job Lifecycle}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=0.6cm and 0.3cm,
  flowstep/.style={draw, rounded corners, minimum width=4cm, minimum height=0.8cm,
               font=\small, align=center},
  planstep/.style={flowstep, fill=purple!10},
  schedstep/.style={flowstep, fill=green!10},
  toolstep/.style={flowstep, fill=orange!10},
  memstep/.style={flowstep, fill=blue!10},
  arrow/.style={-{Stealth[length=6pt]}, thick}
]
  \node[planstep] (submit) {1. Client submits job};
  \node[planstep, below=of submit] (decompose) {2. Decomposer creates DAG};
  \node[planstep, below=of decompose] (orderbook) {3. OrderBook posts demand};
  \node[planstep, below=of orderbook] (match) {4. Agent submits proposal};
  \node[planstep, below=of match] (clear) {5. Market clears, escrow holds};
  \node[schedstep, below=of clear] (assign) {6. Agent assigned via Scheduler};
  \node[toolstep, below=of assign] (tools) {7. Agent invokes tools};
  \node[memstep, below=of tools] (memory) {8. Results persisted to memory};
  \node[schedstep, below=of memory] (eval) {9. Evaluator scores output};
  \node[planstep, below=of eval] (settle) {10. Escrow settles, reputation updates};

  \draw[arrow] (submit) -- (decompose);
  \draw[arrow] (decompose) -- (orderbook);
  \draw[arrow] (orderbook) -- (match);
  \draw[arrow] (match) -- (clear);
  \draw[arrow] (clear) -- (assign);
  \draw[arrow] (assign) -- (tools);
  \draw[arrow] (tools) -- (memory);
  \draw[arrow] (memory) -- (eval);
  \draw[arrow] (eval) -- (settle);

  % Labels for subsystems
  \node[right=0.5cm of submit, font=\scriptsize\color{purple!60}] {Planner};
  \node[right=0.5cm of assign, font=\scriptsize\color{green!60!black}] {Scheduler};
  \node[right=0.5cm of tools, font=\scriptsize\color{orange!60!black}] {Tools};
  \node[right=0.5cm of memory, font=\scriptsize\color{blue!60}] {Memory};
\end{tikzpicture}
\caption{Complete job lifecycle through all four subsystems. Colors indicate which
subsystem is primary at each step.}
\label{fig:lifecycle}
\end{figure}

\subsection{Step-by-Step Walkthrough}

\subsubsection{Step 1--2: Job Submission and Decomposition (Planner Engine)}

\begin{lstlisting}[language=elixir, caption={Job submission and decomposition}]
# Client submits a web testing job
job_spec = %{
  client_id: "client_1",
  task_id: "task_42",
  description: "Full web app test suite",
  required_capabilities: [:playwright, :k6, :reporting],
  budget_ceiling: 5000,
  subtask_specs: [
    %{description: "E2E tests",
      required_capabilities: [:playwright],
      estimated_credits: 2000, estimated_duration: 60},
    %{description: "Load tests",
      required_capabilities: [:k6],
      estimated_credits: 1500, estimated_duration: 45},
    %{description: "Generate report",
      required_capabilities: [:reporting],
      estimated_credits: 500, estimated_duration: 15,
      depends_on: [0, 1]}
  ]
}

# Decompose into DAG with parallel execution levels
{:ok, dag} = PlannerEngine.Decomposer.decompose(job_spec)
schedule = PlannerEngine.Decomposer.topological_sort(dag)
# => [["task_42_sub_0", "task_42_sub_1"], ["task_42_sub_2"]]
\end{lstlisting}

\subsubsection{Step 3--5: Order Book Matching and Market Clearing (Planner Engine)}

\begin{lstlisting}[language=elixir, caption={Market clearing with escrow}]
# Post demand to order book
:ok = PlannerEngine.OrderBook.post_demand(%{
  client_id: "client_1",
  task_id: "task_42",
  required_capabilities: [:playwright, :k6],
  budget_ceiling: 5000
})

# Agent submits proposal
:ok = PlannerEngine.OrderBook.submit_proposal(%{
  agent_id: "agent_web_tester",
  task_id: "task_42",
  execution_plan: "Playwright E2E + k6 load tests",
  estimated_credits: 3500,
  estimated_duration: 120,
  confidence_score: 0.92
})

# Market clears: escrow holds 3500 credits
{:ok, contract} = PlannerEngine.Market.clear_market("task_42")
# contract.escrow_id now references a held escrow record
\end{lstlisting}

\subsubsection{Step 6: Agent Assignment (Scheduler)}

\begin{lstlisting}[language=elixir, caption={Scheduler assigns agent to job}]
# Start agent under supervision
{:ok, _pid} = AgentScheduler.start_agent(
  "agent_web_tester",
  %{name: "WebTester", capabilities: [:playwright, :k6]},
  credits: 0,
  oversight: :autonomous_escalation
)

# Assign the job -- transitions agent to :running
:ok = AgentScheduler.Agent.assign_job(
  "agent_web_tester",
  %{id: "task_42", contract_id: contract.id}
)
\end{lstlisting}

\subsubsection{Step 7--8: Tool Invocation with Memory Persistence}

\begin{lstlisting}[language=elixir, caption={Agent executes steps with tools and memory}]
# Grant tool capabilities for this contract
{:ok, cap} = ToolInterface.grant_capability(
  "agent_web_tester", "code-exec",
  permissions: [:invoke], rate_limit: 60
)

# Freeze tool registry (contract-locked)
ToolInterface.freeze()

# Execute durable step: invoke tool, persist to memory
{:ok, e2e_results} = AgentScheduler.Agent.execute_step(
  "agent_web_tester", "e2e_tests", fn ->
    # Invoke sandboxed tool
    {:ok, result} = ToolInterface.invoke(
      "agent_web_tester", "code-exec",
      %{"code" => playwright_script, "language" => "javascript"},
      cap
    )

    # Persist result as typed memory
    {:ok, _} = MemoryLayer.Memory.create(
      %MemoryLayer.Schema.FactData{
        assertion: "E2E tests passed: 47/50 scenarios",
        confidence: 0.94,
        source: "agent_web_tester/e2e_tests"
      })

    result
  end
)
\end{lstlisting}

If the agent crashes after the tool returns, the memoization store replays
\texttt{"e2e\_tests"} from cache. The memory persists in Mnesia regardless
of agent process state.

\subsubsection{Step 9: Evaluation (Scheduler)}

\begin{lstlisting}[language=elixir, caption={6-dimensional evaluation}]
# Agent completes and submits result
:ok = AgentScheduler.Agent.complete("agent_web_tester", %{
  e2e_results: e2e_results,
  load_results: load_results,
  report: final_report
})

# Evaluator scores the output
{:ok, evaluation} = AgentScheduler.Evaluator.evaluate(
  "agent_web_tester",
  %{
    quality: 0.87,
    adherence: 0.92,
    speed: 0.78,
    cost: 0.85,
    error_rate: 0.06,       # inverted to 0.94
    revision_count: 0.08    # inverted to 0.92
  })
# => %{composite: 0.8771, reputation: 0.8771, ...}
\end{lstlisting}

\subsubsection{Step 10: Settlement and Reputation (Planner Engine)}

\begin{lstlisting}[language=elixir, caption={Escrow settlement and reputation update}]
# Complete contract: settle escrow, distribute revenue
{:ok, %{contract: completed, split: split}} =
  PlannerEngine.Market.complete_contract(
    contract.id,
    [0.87, 0.92, 0.78, 0.85, 0.94, 0.92]
  )

# Revenue split:
# split.operator = 2450 (70% of 3500)
# split.platform = 525  (15%)
# split.llm_reserve = 525 (15%)

# Reputation updated with anti-gaming checks
trust = PlannerEngine.Reputation.trust_score("agent_web_tester")
\end{lstlisting}

\subsection{Pipeline Composition Across Layers}

For multi-stage jobs (like the web testing pipeline), the streaming pipeline
composes all four subsystems across multiple agents:

\begin{lstlisting}[language=elixir, caption={Streaming pipeline across subsystems}]
{:ok, pipe_id} = AgentScheduler.create_pipeline(:web_testing, [
  {:recon,     publishes: [:page_discovered, :api_found,
                           :sitemap_built]},
  {:behavior,  subscribes: [:page_discovered, :sitemap_built],
               publishes: [:test_generated, :flow_mapped]},
  {:load,      subscribes: [:api_found, :flow_mapped],
               publishes: [:load_result, :perf_metric]},
  {:observer,  subscribes: [:test_generated, :load_result,
                            :perf_metric],
               publishes: [:anomaly_detected]},
  {:synthesis, subscribes: [:test_generated, :load_result,
                            :anomaly_detected],
               publishes: [:final_report]}
])
\end{lstlisting}

Each stage is assigned to an agent from the scheduler's pool. The agent
at each stage invokes tools (Tool Interface), persists intermediate results
(Memory Layer), and publishes events that downstream stages consume.
The pipeline constraint---$\text{sub}(s_i) \subseteq \bigcup_{j < i} \text{pub}(s_j)$---ensures
correct data flow.

Streaming reduces latency from $\sum_i t_i$ (sequential) to the critical
path length, typically yielding 40--60\% wall-clock time reduction.


% ========================================================================
\section{Production System: Agent-Hero Marketplace}\label{sec:production}

The composed AgentOS system is deployed as the backend for the Agent-Hero
marketplace, a production platform for hiring AI agent teams.

\subsection{Technology Stack}

\begin{table}[H]
\centering
\caption{Agent-Hero production stack}
\begin{tabular}{ll}
\toprule
\textbf{Component} & \textbf{Technology} \\
\midrule
Frontend           & Next.js 16, React 19, App Router \\
Database           & Supabase (PostgreSQL + Auth + RLS + Realtime) \\
Payments           & Stripe (checkout sessions, webhooks) \\
AI SDK             & Vercel AI SDK v6 (multi-provider) \\
Agent Runtime      & AgentOS (Elixir/OTP on BEAM) \\
Queue              & Inngest (durable execution) \\
Agent Protocol     & MCP (\texttt{StreamableHTTPClientTransport}) \\
Quick Agents       & E2B sandbox (10-min timeout, ephemeral) \\
Long Agents        & Fly.io (hours/days, persistent) \\
Deployment         & Vercel (frontend), Docker/Fly.io (agents) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Mapping AgentOS to Production}

Each AgentOS subsystem maps to a production concern:

\begin{description}[style=nextline]
  \item[Memory Layer $\to$ Supabase PostgreSQL]
    In production, the Memory Layer's Mnesia backend is replaced by
    Supabase PostgreSQL with Row Level Security. The ETS working memory
    layer maps to Redis for distributed caching. The schema type system
    maps to Supabase table schemas with Zod v4 validation on the TypeScript side.

  \item[Tool Interface $\to$ MCP Servers]
    Each agent type connects to external MCP servers via
    \texttt{StreamableHTTPClientTransport}. The 3-tier registry maps to:
    builtin tools (Vercel AI SDK built-in tools),
    sandbox tools (E2B ephemeral environments),
    and MCP tools (agent-specific MCP servers deployed on Fly.io).

  \item[Agent Scheduler $\to$ Inngest + Vercel]
    The durable execution model (memoized steps with replay on crash)
    maps directly to Inngest's step functions. The credit-weighted scheduler
    maps to the Supabase \texttt{profiles.credits} column with atomic RPC
    updates via \texttt{adminClient.rpc()}.

  \item[Planner Engine $\to$ Stripe + Supabase]
    The escrow system maps to Stripe checkout sessions: credits are purchased
    via Stripe, held in the database during execution, and distributed upon
    completion. The order book maps to the job posting and proposal submission
    flow in the Agent-Hero UI. The reputation system maps to the
    \texttt{agent\_ratings} table with EWMA computation.
\end{description}

\subsection{Job Lifecycle in Production}

\begin{enumerate}
  \item \textbf{Client posts job} --- Creates a job record in Supabase via the
        Next.js API route. Triggers decomposition.
  \item \textbf{Planner decomposes} --- Inngest function decomposes the job into
        subtasks and posts them as individual demands.
  \item \textbf{Order book matches} --- Agents (operators) submit proposals.
        The matching engine selects the best proposal per subtask.
  \item \textbf{Market clears} --- Stripe holds the payment. Contract record
        created in Supabase.
  \item \textbf{Scheduler runs agents} --- Inngest dispatches agent execution.
        Each agent connects to its MCP server, invokes tools, persists results.
  \item \textbf{Evaluator scores} --- 6-dimensional quality assessment.
        Client can approve or request revisions.
  \item \textbf{Settlement} --- Stripe releases payment. Revenue split applied.
        Reputation updated.
\end{enumerate}

\subsection{Deployment Architecture}

\begin{lstlisting}[caption={Deployment topology}, basicstyle=\ttfamily\footnotesize]
                    +-------------------+
                    |    Vercel CDN     |
                    |  Next.js Frontend |
                    +---------+---------+
                              |
                    +---------+---------+
                    |   Supabase Cloud  |
                    |  PostgreSQL + Auth|
                    |  + RLS + Realtime |
                    +---------+---------+
                              |
              +---------------+---------------+
              |                               |
    +---------+---------+           +---------+---------+
    |   Inngest Queue   |           |   Stripe Webhooks |
    | (durable steps)   |           |   (payments)      |
    +---------+---------+           +-------------------+
              |
    +---------+---------+
    |   Agent Runtimes  |
    | +----- Fly.io ---+|
    | | MCP Server     ||
    | | AgentOS/BEAM   ||
    | +----------------+|
    | +----- E2B ------+|
    | | Sandbox (10m)  ||
    | +----------------+|
    +-------------------+
\end{lstlisting}


% ========================================================================
\section{System Properties}\label{sec:properties}

We now establish three key properties of the composed system: fault isolation,
type safety, and resource fairness.

\subsection{Fault Isolation}

\begin{theorem}[Subsystem Independence]\label{thm:fault-isolation}
A crash in subsystem $S_i$ does not corrupt the state of subsystem $S_j$ for $i \neq j$.
\end{theorem}

\begin{proof}
Each subsystem runs as a separate OTP application under \texttt{AgentOS.Supervisor}
with \texttt{:one\_for\_one} strategy. Subsystem state is held in GenServer
processes with isolated heaps (the BEAM guarantees no shared mutable state between
processes). When a subsystem crashes:

\begin{enumerate}
  \item The supervisor detects the crash via process monitoring.
  \item Only the crashed subsystem's supervision subtree is restarted.
  \item Other subsystems continue operating with their state intact.
  \item Mnesia tables (used by Memory Layer and Escrow) survive process crashes
        because they are backed by \texttt{disc\_copies}.
\end{enumerate}

The only cross-subsystem dependency at crash time is that a subsystem attempting
to call a crashed subsystem's GenServer will receive \texttt{\{:error, :noproc\}}
(or a timeout), which is handled by pattern matching in all public APIs.
\end{proof}

\begin{corollary}[Cascading Failure Prevention]
The maximum blast radius of any single process crash is the subsystem that process
belongs to, bounded by the subsystem's \texttt{max\_restarts/max\_seconds} configuration.
\end{corollary}

\subsection{Agent-Level Fault Isolation}

Within the Agent Scheduler, individual agent crashes are further isolated:

\begin{theorem}[Agent Independence]\label{thm:agent-isolation}
A crash in agent $A_i$ does not affect agent $A_j$ for $i \neq j$, and the
agent's durable execution guarantees that completed steps are not re-executed
on restart.
\end{theorem}

\begin{proof}
Agents run under \texttt{AgentScheduler.Supervisor} (a DynamicSupervisor with
\texttt{:one\_for\_one} strategy). Each agent is a GenServer with:

\begin{itemize}
  \item Its own heap (BEAM process isolation).
  \item A unique registration in \texttt{AgentScheduler.Registry}.
  \item A \texttt{:transient} restart policy (only restarted on abnormal exit).
  \item A 30-second shutdown timeout for graceful checkpointing.
\end{itemize}

For durable execution: the \texttt{memo\_store} maps step IDs to cached results.
On crash and restart, the memo store is lost (it was in-process state), but
if the agent persisted results to the Memory Layer (as shown in
Section~\ref{subsec:agent-memory}), those results survive in Mnesia and can
be used to skip completed steps.
\end{proof}

\subsection{Type Safety Across Boundaries}

\begin{theorem}[Cross-Module Type Safety]\label{thm:type-safety}
The composition of subsystems preserves type safety: if each subsystem's internal
invariants hold, the composed system's invariants hold.
\end{theorem}

\begin{proof}[Proof sketch]
Type safety is enforced at three levels:

\begin{enumerate}
  \item \textbf{Compile-time (Dialyzer):} Each module declares \texttt{@spec} type
        annotations. Dialyzer performs success typing across module boundaries,
        catching type mismatches at compile time.

  \item \textbf{Runtime (Struct enforcement):} Memory schemas use
        \texttt{@enforce\_keys} to guarantee required fields are present.
        Capability tokens enforce their structure via the \texttt{@enforce\_keys}
        attribute. Pattern matching in GenServer callbacks rejects malformed messages.

  \item \textbf{Protocol (Message contracts):} Cross-subsystem communication uses
        typed message tuples: \texttt{\{:execute\_step, step\_id, fun\}},
        \texttt{\{:hold, client\_id, amount, contract\_id\}}, etc. Each GenServer's
        \texttt{handle\_call} pattern-matches on the message structure, rejecting
        messages that don't conform.
\end{enumerate}

Since message passing is the only communication mechanism between subsystems
(no shared mutable state), type safety at the message boundary is sufficient
to guarantee cross-module type safety.
\end{proof}

\subsection{Resource Fairness}

\begin{theorem}[Credit-Weighted Fairness]\label{thm:fairness}
The scheduler's avruntime mechanism ensures that no client can monopolize agent
resources: a client's scheduling priority decreases proportionally to their
resource consumption relative to their credit balance.
\end{theorem}

\begin{proof}
From Equation~\ref{eq:avruntime}, after consuming $k$ agent invocations, client $c$'s
avruntime is:
\[
  \text{avruntime}_k(c) = \sum_{i=1}^{k} \frac{\kappa_{A_i} \cdot \text{cr}_0}{\text{cr}_c(i) \cdot w(p_c)}
\]

Since $\text{cr}_c(i)$ decreases with each invocation (credits are consumed), the
denominator shrinks and avruntime grows faster. Meanwhile, a client $c'$ with more
remaining credits has $\text{cr}_{c'}(i) > \text{cr}_c(i)$, so their avruntime grows
slower. The scheduler always dispatches the minimum-avruntime client, ensuring
$c'$ is prioritized.

The contracted/marketplace two-tier system provides a $4\times$ weight to contracted
clients, but within each tier, the fairness guarantee holds.
\end{proof}

\subsection{Financial Conservation}

\begin{theorem}[Credit Conservation]\label{thm:conservation}
The total credits in the system is invariant under all escrow operations:
\[
  \sum_p \text{available}(p) + \sum_e \text{held}(e) + \sum_d \text{distributed}(d) = C_0
\]
where $C_0$ is the total credits ever purchased.
\end{theorem}

\begin{proof}
Each escrow operation is a Mnesia transaction that conserves the total:

\begin{itemize}
  \item \texttt{hold}: decrements available by $n$, increments held by $n$.
  \item \texttt{settle(:release)}: decrements held by $n$, increments distributed by $n$.
  \item \texttt{settle(:refund)}: decrements held by $n$, increments available by $n$.
\end{itemize}

Mnesia transactions are serialized, preventing race conditions. The balance check
(\texttt{available >= amount}) in \texttt{hold} prevents negative balances.
No operation creates or destroys credits.
\end{proof}


% ========================================================================
\section{Implementation Summary}\label{sec:implementation}

\subsection{Source Code Structure}

\begin{table}[H]
\centering
\caption{Source code organization}
\begin{tabular}{llr}
\toprule
\textbf{Application} & \textbf{Modules} & \textbf{Lines} \\
\midrule
\texttt{agent\_os/}        & \texttt{AgentOS}, \texttt{Application}  & $\sim$80 \\
\texttt{agent\_scheduler/} & \texttt{Agent}, \texttt{Scheduler}, \texttt{Pipeline},
                              \texttt{Evaluator}, \texttt{Supervisor}  & $\sim$900 \\
\texttt{tool\_interface/}  & \texttt{Registry}, \texttt{Capability}, \texttt{Sandbox},
                              \texttt{MCPClient}, \texttt{Audit}  & $\sim$800 \\
\texttt{memory\_layer/}    & \texttt{Memory}, \texttt{Schema}, \texttt{Storage},
                              \texttt{Graph}, \texttt{Version}  & $\sim$900 \\
\texttt{planner\_engine/}  & \texttt{OrderBook}, \texttt{Escrow}, \texttt{Market},
                              \texttt{Decomposer}, \texttt{Reputation}  & $\sim$900 \\
\midrule
\textbf{Total}             & \textbf{24 modules} & $\sim$\textbf{3,500} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Key Design Patterns}

\begin{description}[style=nextline]
  \item[GenServer for Stateful Services]
    Every stateful component (Registry, Scheduler, Evaluator, Pipeline, OrderBook,
    Escrow, Market, Reputation, Storage, Graph, Audit, Schema Registry) is a
    GenServer. This provides: serialized state access (no race conditions),
    process isolation (crash containment), and a standard callback interface
    (\texttt{init}, \texttt{handle\_call}, \texttt{handle\_cast}, \texttt{handle\_info}).

  \item[DynamicSupervisor for Agent Pools]
    Agents are created dynamically when jobs are assigned and terminated when
    jobs complete. The DynamicSupervisor provides on-demand process creation
    with automatic restart on crash.

  \item[Registry for O(1) Lookup]
    Agent processes are registered in \texttt{AgentScheduler.Registry} (an
    Elixir \texttt{Registry} with \texttt{:unique} keys) for constant-time
    lookup by agent ID. This avoids the need for a centralized process table.

  \item[Behaviour Callbacks for Polymorphism]
    The \texttt{MemoryLayer.StorageBackend} behaviour defines the interface
    that all storage backends must implement: \texttt{save}, \texttt{recall},
    \texttt{search}, \texttt{delete}, \texttt{update}. Adding a new backend
    (ChromaDB, FalkorDB, PostgreSQL) requires implementing these five callbacks.

  \item[Telemetry for Observability]
    Every significant event emits a \texttt{:telemetry} event:
    \texttt{[:agent\_scheduler, :agent, :step\_completed]},
    \texttt{[:tool\_interface, :invocation]}, etc. These can be consumed by
    monitoring systems (Prometheus, Datadog, CloudWatch) without modifying
    application code.

  \item[Mnesia for ACID Transactions]
    Financial operations (escrow hold/settle/refund, balance updates) use
    Mnesia transactions for atomicity and isolation. Mnesia is started during
    the MemoryLayer initialization phase, ensuring it is available before
    the Escrow system starts.
\end{description}


% ========================================================================
\section{Composition Algebra}\label{sec:algebra}

We can formalize the composition structure of AgentOS using a simple algebra
of module interfaces.

\subsection{Module Interfaces as Types}

Each subsystem exposes a set of typed functions (its public API).
We write $\mathcal{I}(S)$ for the interface of subsystem $S$:

\begin{align*}
  \mathcal{I}(\text{Memory}) &= \{\text{create}, \text{data}, \text{evolve}, \text{search}, \text{link}, \text{traverse}\} \\
  \mathcal{I}(\text{Tools})  &= \{\text{invoke}, \text{grant}, \text{discover}, \text{freeze}, \text{list}\} \\
  \mathcal{I}(\text{Sched})  &= \{\text{submit}, \text{start\_agent}, \text{pipeline}, \text{evaluate}\} \\
  \mathcal{I}(\text{Plan})   &= \{\text{post}, \text{propose}, \text{clear}, \text{hold}, \text{settle}, \text{decompose}\}
\end{align*}

\subsection{Composition as Interface Products}

Pairwise composition yields new capabilities that use functions from both interfaces:

\begin{align*}
  \mathcal{I}(\text{Sched}) \times \mathcal{I}(\text{Tools}) &\to \text{durable\_tool\_invocation} \\
  \mathcal{I}(\text{Sched}) \times \mathcal{I}(\text{Memory}) &\to \text{persistent\_agent\_state} \\
  \mathcal{I}(\text{Tools}) \times \mathcal{I}(\text{Memory}) &\to \text{cached\_tool\_results} \\
  \mathcal{I}(\text{Plan}) \times \mathcal{I}(\text{Sched}) &\to \text{market\_dispatched\_agents} \\
  \mathcal{I}(\text{Plan}) \times \mathcal{I}(\text{Memory}) &\to \text{persistent\_market\_state} \\
  \mathcal{I}(\text{Plan}) \times \mathcal{I}(\text{Tools}) &\to \text{capability\_verified\_matching}
\end{align*}

\subsection{The Full Product}

The AgentOS umbrella is the product of all four interfaces:
\[
  \mathcal{I}(\text{AgentOS}) = \mathcal{I}(\text{Memory}) \times \mathcal{I}(\text{Tools}) \times \mathcal{I}(\text{Sched}) \times \mathcal{I}(\text{Plan})
\]

with the startup ordering constraint:
\[
  \text{Memory} \prec \text{Tools} \prec \text{Sched} \prec \text{Plan}
\]

where $A \prec B$ means ``$A$ must be fully initialized before $B$ starts.''

The facade module $\texttt{AgentOS}$ projects from this product to a simplified
interface:
\[
  \pi : \mathcal{I}(\text{AgentOS}) \to \{\text{start}, \text{submit\_job}, \text{status}\}
\]


% ========================================================================
\section{Emergent Properties of Composition}\label{sec:emergent}

Beyond the pairwise compositions analyzed in Section~\ref{sec:pairwise},
the full four-way composition exhibits emergent properties that require
all four subsystems to interact.

\subsection{End-to-End Durable Execution}

The combination of:
\begin{itemize}
  \item Agent memoization (Scheduler)
  \item Tool sandboxing (Tool Interface)
  \item Mnesia persistence (Memory Layer)
  \item Escrow transactions (Planner Engine)
\end{itemize}
yields end-to-end durable execution: a job can crash at any point and resume
from the last successful step without re-executing tools, losing memory, or
double-spending credits.

\subsection{Self-Improving Reputation}

The feedback loop:
\[
  \text{Evaluation} \xrightarrow{\text{scores}} \text{Reputation}
  \xrightarrow{\text{trust}} \text{OrderBook}
  \xrightarrow{\text{matching}} \text{Contracts}
  \xrightarrow{\text{execution}} \text{Evaluation}
\]
creates a self-improving system where high-performing agents receive more
contracts (via lower cost functionals), which produces more evaluations,
which refines their reputation scores. The anti-gaming detectors prevent
this feedback loop from being exploited.

\subsection{Adaptive Pipeline Composition}

The combination of DAG decomposition (Planner) with streaming pipelines
(Scheduler) enables adaptive execution: if an agent at stage $i$ discovers
new information (persisted in Memory), the planner can dynamically add new
subtasks to the DAG and extend the pipeline.


% ========================================================================
\section{Related Work}\label{sec:related}

\subsection{Agent Frameworks}

\textbf{LangChain}~\citep{langchain2023} and \textbf{AutoGPT}~\citep{autogpt2023}
provide agent orchestration but lack formal process management, financial
settlement, and typed memory systems. \textbf{CrewAI}~\citep{crewai2024}
introduces team-based agent coordination but without market mechanisms
or capability-based security.

\subsection{Operating System Analogies}

\textbf{AIOS}~\citep{mei2024aios} proposes an LLM-based OS with agent scheduling
and tool management but does not address financial mechanisms or typed memory.
\textbf{AgentOS} by SWE-agent~\citep{yang2024swe} focuses on software engineering
tasks without general-purpose memory or market orchestration.

\subsection{OTP and Supervision}

The Erlang/OTP supervision tree model~\citep{armstrong2003thesis} has been
applied to distributed systems for decades. Our contribution is applying
these patterns specifically to AI agent management, where the unique challenges
include durable execution across LLM calls, capability-based tool access,
typed memory for agent cognition, and market-based resource allocation.

\subsection{Market Mechanisms for AI}

\textbf{MIRI}~\citep{soares2015corrigibility} and
\textbf{Anthropic}~\citep{bai2022constitutional} study AI alignment but not
market-based orchestration. The closest work is economic mechanism design
for multi-agent systems~\citep{shoham2009multiagent}, which we instantiate
with concrete OTP implementations and Stripe-backed financial settlement.


% ========================================================================
\section{Conclusion}\label{sec:conclusion}

This paper has presented the synthesis of four independently developed subsystems
into a unified AI operating system. The key contributions are:

\begin{enumerate}
  \item \textbf{Modular composition via OTP supervision}: The \texttt{AgentOS}
        umbrella application composes Memory Layer, Tool Interface, Agent Scheduler,
        and Planner Engine with a dependency-ordered supervision tree that provides
        fault isolation and automatic recovery.

  \item \textbf{Six pairwise emergent capabilities}: Every pair of subsystems
        yields capabilities that neither provides alone: durable tool invocation,
        persistent agent state, cached tool results, market-dispatched agents,
        persistent market state, and capability-verified matching.

  \item \textbf{Complete production job lifecycle}: The composed system supports
        the full cycle from job submission through decomposition, market matching,
        agent execution with tools and memory, evaluation, and financial settlement.

  \item \textbf{Formal guarantees}: We proved fault isolation (Theorem~\ref{thm:fault-isolation}),
        agent independence (Theorem~\ref{thm:agent-isolation}), cross-module type
        safety (Theorem~\ref{thm:type-safety}), credit-weighted fairness
        (Theorem~\ref{thm:fairness}), and credit conservation
        (Theorem~\ref{thm:conservation}).

  \item \textbf{Production validation}: The system is deployed as the Agent-Hero
        marketplace, a Next.js + Supabase + Stripe platform running on BEAM.
\end{enumerate}

The central lesson is that complex AI systems need not be monolithic.
By designing each subsystem as a self-contained module with a well-defined
interface, and composing them through standard functional programming patterns
(supervision trees, message passing, typed interfaces), we achieve both
modularity and emergent capability. The BEAM virtual machine provides the
ideal substrate: lightweight processes for massive agent concurrency,
supervision trees for fault tolerance, and Mnesia for distributed
transactional state.

\subsection{Future Work}

\begin{itemize}
  \item \textbf{Distributed deployment}: Extending AgentOS across multiple
        BEAM nodes for horizontal scaling, using Erlang distribution for
        cross-node GenServer calls and Mnesia replication for distributed memory.

  \item \textbf{Formal verification}: Using session types or
        Scribble~\citep{honda2012scribble} to formally verify the message-passing
        protocols between subsystems.

  \item \textbf{Adaptive scheduling}: Replacing the static avruntime formula with
        a learned scheduling policy that adapts to workload characteristics.

  \item \textbf{Vector memory backends}: Integrating ChromaDB for semantic search
        and FalkorDB for graph traversal, extending the Memory Layer's storage
        router with production-grade backends.

  \item \textbf{Multi-agent negotiation}: Extending the order book with
        multi-round negotiation protocols where agents can counter-propose
        and clients can revise requirements.
\end{itemize}


% ========================================================================
\appendix
\section{Module Dependency Matrix}\label{app:deps}

Table~\ref{tab:deps} shows which modules each subsystem calls, establishing
the complete dependency graph of the composed system.

\begin{table}[H]
\centering
\caption{Cross-subsystem module dependencies. Rows are callers, columns are callees.}
\label{tab:deps}
\small
\begin{tabular}{l|cccc}
\toprule
 & \textbf{Memory} & \textbf{Tools} & \textbf{Scheduler} & \textbf{Planner} \\
\midrule
\textbf{Memory}    & ---     & ---     & ---       & --- \\
\textbf{Tools}     & read    & ---     & ---       & --- \\
\textbf{Scheduler} & persist & invoke  & ---       & --- \\
\textbf{Planner}   & persist & verify  & dispatch  & --- \\
\textbf{AgentOS}   & status  & list    & ---       & submit \\
\bottomrule
\end{tabular}
\end{table}

The matrix is lower-triangular (with the exception of AgentOS, which calls
all four), confirming that the dependency order
Memory $\prec$ Tools $\prec$ Scheduler $\prec$ Planner is acyclic.


\section{Supervision Tree Diagram}\label{app:supervision}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  every node/.style={font=\ttfamily\footnotesize},
  level 1/.style={sibling distance=4cm, level distance=1.8cm},
  level 2/.style={sibling distance=2.5cm, level distance=1.8cm},
  level 3/.style={sibling distance=1.5cm, level distance=1.5cm},
  sup/.style={draw, rounded corners, fill=red!10, minimum width=2.2cm, align=center},
  gen/.style={draw, rounded corners, fill=blue!10, minimum width=2cm, align=center},
  dyn/.style={draw, rounded corners, fill=green!10, minimum width=2cm, align=center},
  edge from parent/.style={draw, -{Stealth[length=5pt]}},
]
  \node[sup] {AgentOS\\Supervisor}
    child { node[sup] {MemoryLayer\\Supervisor}
      child { node[gen] {Schema\\Registry} }
      child { node[gen] {Storage\\Router} }
      child { node[gen] {Graph\\Manager} }
    }
    child { node[sup] {ToolInterface\\Supervisor}
      child { node[gen] {Tool\\Registry} }
      child { node[gen] {Audit\\Logger} }
    }
    child { node[sup] {AgentScheduler\\AppSupervisor}
      child { node[gen] {Evaluator} }
      child { node[gen] {Scheduler} }
      child { node[gen] {Pipeline} }
      child { node[dyn] {Agent\\Pool} }
    }
    child { node[sup] {PlannerEngine\\Supervisor}
      child { node[gen] {Escrow} }
      child { node[gen] {OrderBook} }
      child { node[gen] {Reputation} }
      child { node[gen] {Market} }
    };
\end{tikzpicture}
\caption{Complete supervision tree of AgentOS. Red = supervisors, blue = GenServers,
green = DynamicSupervisor.}
\label{fig:full-tree}
\end{figure}


\section{Message Flow Diagram}\label{app:messages}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=3cm,
  box/.style={draw, rounded corners, minimum width=2.5cm, minimum height=1cm,
              font=\small\ttfamily, align=center},
  arrow/.style={-{Stealth[length=6pt]}, thick},
  label/.style={font=\scriptsize, midway, above, sloped}
]
  \node[box, fill=purple!10] (client) {Client};
  \node[box, fill=purple!10, right=of client] (planner) {Planner\\Engine};
  \node[box, fill=green!10, below=2cm of planner] (scheduler) {Agent\\Scheduler};
  \node[box, fill=orange!10, right=of scheduler] (tools) {Tool\\Interface};
  \node[box, fill=blue!10, below=2cm of scheduler] (memory) {Memory\\Layer};

  \draw[arrow] (client) -- (planner) node[label] {submit\_job};
  \draw[arrow] (planner) -- (scheduler) node[label] {assign\_job};
  \draw[arrow] (scheduler) -- (tools) node[label] {invoke};
  \draw[arrow] (scheduler) -- (memory) node[label] {persist};
  \draw[arrow] (tools) -- (memory) node[label] {cache};
  \draw[arrow, dashed] (scheduler) to[bend right=30] node[label, below, sloped] {evaluate} (planner);
\end{tikzpicture}
\caption{Primary message flows between subsystems during job execution.
Solid arrows are synchronous \texttt{GenServer.call}; dashed arrows are
asynchronous feedback loops.}
\label{fig:messages}
\end{figure}


\section{Algorithm: Complete Job Execution}\label{app:algorithm}

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\KwIn{Job specification $J$ with client $c$, task $\tau$, subtasks $[s_1, \ldots, s_n]$, budget $B$}
\KwOut{Evaluation result and financial settlement}
\BlankLine
\tcp{Phase 1: Decomposition (Planner Engine)}
$\text{DAG} \gets \texttt{Decomposer.decompose}(J)$\;
$\text{levels} \gets \texttt{Decomposer.topological\_sort}(\text{DAG})$\;
\BlankLine
\tcp{Phase 2: Market Matching (Planner Engine)}
\texttt{OrderBook.post\_demand}$(c, \tau, B)$\;
\ForEach{proposal $p$ from agents}{
  \texttt{OrderBook.submit\_proposal}$(p)$\;
}
$\text{contract} \gets \texttt{Market.clear\_market}(\tau)$\;
$\text{escrow\_id} \gets \texttt{Escrow.hold}(c, p.\text{credits}, \text{contract.id})$\;
\BlankLine
\tcp{Phase 3: Execution (Scheduler + Tools + Memory)}
$\text{agent} \gets \texttt{Scheduler.start\_agent}(\text{contract.operator\_id})$\;
$\text{cap} \gets \texttt{ToolInterface.grant\_capability}(\text{agent}, \text{tools})$\;
\texttt{ToolInterface.freeze()}\;
\ForEach{level $L$ in levels}{
  \ForEach{subtask $s$ in $L$ \textbf{in parallel}}{
    $\text{result} \gets \texttt{Agent.execute\_step}(\text{agent}, s.\text{id}, \lambda.\;$
    $\quad \texttt{ToolInterface.invoke}(\text{agent}, s.\text{tool}, s.\text{input}, \text{cap})$
    $)$\;
    \texttt{Memory.create}$(\text{FactData}\{s.\text{id}, \text{result}\})$\;
  }
}
\texttt{Agent.complete}$(\text{agent}, \text{results})$\;
\BlankLine
\tcp{Phase 4: Evaluation and Settlement}
$\text{scores} \gets \texttt{Evaluator.evaluate}(\text{agent}, \text{quality\_vector})$\;
$\text{split} \gets \texttt{Market.complete\_contract}(\text{contract.id}, \text{quality\_vector})$\;
\texttt{Reputation.record\_quality}$(\text{agent}, \text{quality\_vector})$\;
\Return{$(\text{scores}, \text{split})$}
\caption{Complete job execution through all four AgentOS subsystems}
\label{alg:full-job}
\end{algorithm}


% ========================================================================
\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem[Armstrong(2003)]{armstrong2003thesis}
J.~Armstrong.
\newblock Making reliable distributed systems in the presence of software errors.
\newblock PhD thesis, Royal Institute of Technology, Stockholm, 2003.

\bibitem[AutoGPT(2023)]{autogpt2023}
T.~Richards et~al.
\newblock {AutoGPT}: An autonomous GPT-4 agent.
\newblock \url{https://github.com/Significant-Gravitas/AutoGPT}, 2023.

\bibitem[Bai et~al.(2022)]{bai2022constitutional}
Y.~Bai, S.~Kadavath, S.~Kundu, et~al.
\newblock Constitutional {AI}: Harmlessness from {AI} feedback.
\newblock \emph{arXiv preprint arXiv:2212.08073}, 2022.

\bibitem[CrewAI(2024)]{crewai2024}
J.~Moura.
\newblock {CrewAI}: Framework for orchestrating role-playing AI agents.
\newblock \url{https://github.com/joaomdmoura/crewAI}, 2024.

\bibitem[Honda et~al.(2012)]{honda2012scribble}
K.~Honda, N.~Yoshida, and M.~Carbone.
\newblock Multiparty asynchronous session types.
\newblock \emph{Journal of the ACM}, 63(1):1--67, 2016.

\bibitem[LangChain(2023)]{langchain2023}
H.~Chase.
\newblock {LangChain}: Building applications with {LLMs} through composability.
\newblock \url{https://github.com/langchain-ai/langchain}, 2023.

\bibitem[Mei et~al.(2024)]{mei2024aios}
K.~Mei, Z.~Li, S.~Xu, et~al.
\newblock {AIOS}: {LLM} agent operating system.
\newblock \emph{arXiv preprint arXiv:2403.16971}, 2024.

\bibitem[Shoham and Leyton-Brown(2009)]{shoham2009multiagent}
Y.~Shoham and K.~Leyton-Brown.
\newblock \emph{Multiagent Systems: Algorithmic, Game-Theoretic, and Logical
  Foundations}.
\newblock Cambridge University Press, 2009.

\bibitem[Soares et~al.(2015)]{soares2015corrigibility}
N.~Soares, B.~Fallenstein, E.~Yudkowsky, and S.~Armstrong.
\newblock Corrigibility.
\newblock In \emph{AAAI Workshop on AI and Ethics}, 2015.

\bibitem[Yang et~al.(2024)]{yang2024swe}
J.~Yang, C.~Jimenez, A.~Wettig, et~al.
\newblock {SWE-agent}: Agent-computer interfaces enable automated software engineering.
\newblock \emph{arXiv preprint arXiv:2405.15793}, 2024.

\end{thebibliography}

\end{document}
